---
title: "Covid19 Analysis"
author: "TCJ"
date: "`r Sys.Date()`"
output: pdf_document
---

# Importing Datasets


## During the production of this report, I encounted several technical challenges, some of which I was able to handle, but others required changing or adding chunks of codes in order to successfully knit this report. In addition to the technical challenges, I also noticed that the datasets were updated/changed.

### At the time of this report, I was unable to get the direct link for the csv files, so I had to manually download and import them. The original link can be found below.

```{r, echo=FALSE, include=FALSE}
setwd("/Users/tauancosta/Desktop/Education/Masters - Colorado University Boulder/Data Science as a Field/Week3_exercise_Covid_Data/Covid Analysis")
```
```{r, warning=FALSE, message=FALSE}
library(httr)
library(tidyverse)
library(ggplot2)
library(tidyr)

# Use this link for the original datasets:
#https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series


US_cases <- read_csv("time_series_covid19_confirmed_US.csv")
global_cases <- read_csv("time_series_covid19_deaths_US.csv") 
global_deaths <- read_csv("time_series_covid19_deaths_global.csv") 
US_deaths <- read_csv("time_series_covid19_deaths_US.csv")
```

# Cleaning the Data

## For some reason, while downloading the file, the dates column came with an X. In case that happens, you can use the code below to change to remove the X and substituting the points to slashes global confirmed cases data:
## (for (col in 5:ncol(global_cases)){colnames(global_cases)[col] <- sub("X","", x =colnames(global_cases)[col])})
## (for (col in 1:ncol(global_cases)){colnames(global_cases)[col] <- gsub("\\.","/", x =colnames(global_cases)[col])})


```{r, warning=FALSE, message=FALSE}
#Global Cases

#Let's tidy and  transform the data

#removing columns lat and long
global_cases <- global_cases %>% select(-c(Lat, Long_))%>%
  select(-c(UID, iso2,iso3, code3, FIPS,Admin2, Combined_Key))

# pivoting the global cases
global_cases <- pivot_longer(global_cases, cols = -c("Province_State", "Country_Region"),
                             names_to = "date",values_to = "cases")

#For some reason, while downloading the file, the dates column came with an X. 
#In case that happens, you can use the code below to change it.
#Removing the X and substituting the points to slashes for global deaths data
#(for (col in 5:ncol(global_deaths)){colnames(global_deaths)[col] <- sub("X"
#,"", x =colnames(global_deaths)[col] )})
#(for (col in 1:ncol(global_deaths)){colnames(global_deaths)[col] <- gsub("\\.","/", x =colnames(global_deaths)[col])})

# Cleaning the Global Deaths
global_deaths <- global_deaths %>%
  pivot_longer(cols = -c("Province/State",
 "Country/Region", Lat, Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat, Long))

```
# Joining the datasets 


```{r, eval=FALSE, warning=FALSE, message=FALSE}
global <- global_cases %>%
  full_join(global_deaths) %>%
  mutate(date = mdy(date))

#renaming columns 
global <- global %>% rename(Country_Region = "Country/Region",
          Province_State = "Province/State") 

#removing the ones who have no cases, because they are not relevant for our case
global <- global %>% filter(cases > 0)

```





# US Cases and Deaths

```{r, warning=FALSE, message=FALSE}
#Removing the X and substituting the points to slashes for global deaths data
#(for (col in 5:ncol(US_cases)){colnames(US_cases)[col] <- sub("X","", x =colnames(US_cases)[col] )})
#(for (col in 1:ncol(US_cases)){colnames(US_cases)[col] <- gsub("\\.","/", x =colnames(US_cases)[col])})

US_cases <- US_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
US_cases <- US_cases %>% select(Admin2:cases)



#deaths
#Removing the X and substituting the points to slashes for global deaths data
#(for (col in 5:ncol(US_deaths)){colnames(US_deaths)[col] <- sub("X"
#,"", x =colnames(US_deaths)[col] )})
#(for (col in 1:ncol(US_deaths)){colnames(US_deaths)[col] <- gsub("\\."
#,"/", x =colnames(US_deaths)[col])})

US_deaths <- US_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to ="deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

US <- US_cases %>%
  full_join(US_deaths)
```


# Global Dataset

```{r, eval= FALSE, warning=FALSE, message=FALSE}

global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = "",
        na.rm = TRUE,
        remove = FALSE)

```


## Joining/Merging the data. Unfortunately, I wasn't able to reproduce the joining function used in class - for some unknown reason, R could not locate the "Population" column in either datasets, so I decided to use the merge function.

## Additionally, at the time of this report, something must have changed in the datasets, because I couldn't get the same result. For example, the country Afghanistan did not come with a population count, as it was presented during class. I couldn't identify the reason, so I decided to go on with the report, regardless.
```{r, eval=FALSE, warning=FALSE, message=FALSE}

# URL of the raw CSV file on GitHub
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

# Download the CSV file
response <- GET(uid_lookup_url)

uid <-  read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

```

```{r, eval=FALSE, warning=FALSE, message=FALSE}
merged_data <- merge(global, uid, by = c("Province_State", "Country_Region"), all.x = TRUE)
merged_data <- filter(merged_data, Population.x > 0 )
#Dropping unwanted columns. I decided to drop the Population.y column, since it's just a duplicate of  Population.x
merged_data <- merged_data %>%
  select(-c(UID, FIPS, Population.y))
merged_data
```

# Visualization

```{r, warning=FALSE, message=FALSE}
#grouping by state
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>% ungroup()

#Grouping the US to get the total sum

US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths) ,
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>% ungroup()


#Ploting the results


US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes (y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme (legend.position="bottom",
         axis.text.x = element_text(angle = 90)) +
  labs (title = "COVID19 in US", y= NULL)

```


```{r, warning=FALSE, message=FALSE}
#Plotting the NY state data
state <- "New York"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes (color = "cases")) +
  geom_line(aes (y = deaths, color = "deaths")) +
  geom_point(aes (y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme (legend.position="bottom",
         axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y= NULL)



US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))



US_totals %>%
  ggplot (aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes (y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom",
         axis.text.x = element_text(angle = 90)) +
  labs(title = "Covid19 in US", y= NULL)
```


# Let's see California


```{r, warning=FALSE, message=FALSE}
US_by_state %>% filter(Province_State == "California") %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme (legend.position="bottom",
         axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", "California"), y= NULL)
  
```


# Now let's check the cases per state


```{r, warning=FALSE, message=FALSE}
US_by_state  %>%
  ggplot(aes(x = Province_State, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme (legend.position="bottom",
         axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", "California"), y= NULL)

#let's get the maximum number of cases per state

US_state_totals <- US_by_state %>% group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000* cases /population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)


US_state_totals %>%
  ggplot(aes(x = Province_State, y = cases_per_thou)) +
  geom_line(aes(color = "cases_per_thou")) +
  geom_point(aes(color = "cases_per_thou")) +
  geom_line(aes(y = deaths_per_thou, color = "deaths_per_thou")) +
  geom_point(aes(y = deaths_per_thou, color = "deaths_per_thou")) +
  scale_y_log10() +
  theme (legend.position="bottom",
         axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", "the US"), y= NULL)

```

# Modeling the data


```{r, warning=FALSE, message=FALSE}

mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)

US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))

US_tot_w_pred %>% ggplot +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```

